### 2020-07-12 paper/sex_selection_spacing-ver4.tex
Clean up remaining spell discussions
- use "imposed censoring" rather than generic end of spell
- update interval dummy numbers to reflect birth interval rather than
pregnancy intervals/spell lengths.
- minor edits/clarifications

To find the expected average birth interval, I first calculate, for each woman, 
the probability of giving birth in each period $t$, which I use as weights to calculated 
her expected spell lengths. 
I then average the individual expected spell lengths across women using the parity 
progression probabilities as weights.
Finally, I add nine months to account for spell start to arrive at birth interval.

The predicted parity progression probability is the likelihood of giving birth by the
end of the spell.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is to bias downward the differences in parity progression probabilities.
}
The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over, calculated as shown above.
For comparison, I also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

@@

In the latest period, the shortest average birth interval is 33 months for
women with no education in the second spell, an increase of a couple of
months from the first period.
The longest average interval was 51 months for the third spell for urban women with twelve or 
more years of education who had no boys, which is an increase of approximately a year.
For comparison, the median birth intervals reported for the NFHS have barely
moved over time, staying at approximately 31 months.%
\footnote{
See
\citet[][p.\ 110--112]{International-Institute-for-Population-Sciences-IIPS1995},
\citet[][p.\ 98--103]{International-Institute-for-Population-Sciences-IIPS2000},
\citet[][p.\ 88--91]{International-Institute-for-Population-Sciences-IIPS2007}, and
\citet[][p.\ 81--82]{International-Institute-for-Population-Sciences-IIPS2017}.
Median estimates are in
Appendix Tables \ref{tab:p25_p50_p75_low} through \ref{tab:p25_p50_p75_highest}.
}
This lack of change underscores the importance of accounting for censoring of birth spells 
when trying to understand how birth spacing is changing over time.

Parity progression probabilities are lower the higher the education and are
falling across all groups over time, but the extent of the decline depends crucially on 
the sex composition of prior children and the parity.
Women with no sons still are the most likely to 
have the next birth, and this group has seen often substantially smaller declines in
the likelihood of having another birth than women with one or more sons.%
\footnote{
There is, however, also some evidence that parents prefer a mix of boys and girls: for 
the fourth spell women with two boys \emph{and} a girl
consistently have the lowest parity progression probability.
}
For urban women with eight to eleven years of education, for example, the second spell
probability of a next child declined by eight percentage points if the first child was a 
girl, but 15 percentage points if it was a boy.
Even more extreme is the fourth spell, where the likelihood of a birth declined by six 
percentage points to 75 percent if the woman had three children, but by almost 40 
percentage points to 25 percent if she had two boys and a girl.

@@

To better understand what factors are behind the changes in average spacing, I next 
examine the distribution of birth spacing---represented by the 25th and 75th percentiles 
of predicted birth intervals. 
This approach can capture differences across groups with similar changes in average spacing. 
For example, imagine that sex selection is behind an increase in average birth spacing, 
but that women still try to conceive early. 
In that case, we should see little change in the 25th percentile, but the 75th percentile 
spacing becomes much longer because those who use sex selection will have two or more 
pregnancies rather than one. 
In contrast, a general increase in spacing would show up as lengthening of both percentiles.

@@

I include the spell from marriage to first birth, despite the problems
capturing the exact timing of marriages since the estimated progression
probabilities should not be affected by this problem.
To ensure that the predictions depend solely on sample composition and the estimate, I
use the age at marriage for each woman and predict the likelihood of progressing to
each parity, assuming three years increases in age between each parity.
Shorter assumed duration between births leads to slightly higher predicted 
fertility.

@@

Despite the declines in the predicted cohort fertility, it is still mostly above the 
generally accepted replacement level of 2.1.
Only for urban women with twelve or more years of education is the predicted cohort
fertility clearly below 2.1, and even then, it is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low,
especially for the lower education groups, since it relies only on the
first four births and because the model ignores births that took place past the end
of the 96 months long spell.

@@

The dependent variable is whether the child died within the first twelve
months of life.
The main set of explanatory variables consists of dummies for birth spacing
covering 12 months intervals, starting from nine months after the prior birth 
as above, until 48 months, which includes all births until 96 months.
I use dummies for sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of index child, and the sex composition dummies
are all interacted.
Since the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

I estimate the probability of infant mortality using a logit model.
Figures \ref{fig:mortality_low_med} and \ref{fig:mortality_high_highest} present 
the predicted probability of the second child dying within the first year at the 
possible combinations of index child sex, sex composition of prior children, and 
the birth spacing dummied.%
\footnote{
The online Appendix shows the corresponding graphs for the third child.
}
All other variables take their average values.
The graphs do not show confidence intervals to ease their legibility.

An important caveat is that the estimations do not 
address potential selection problems.
For example, if women who have difficulties conceiving or carrying a pregnancy 
to term also have a higher mortality risk for their offspring, a spurious correlation 
between long birth spacing and mortality may arise \citep{Kozuki2013}. 
Unfortunately, methods used previously to address selection, such
as family fixed effects, do not work well when the number of births is
low as it is for better-educated women \citep{Kozuki2013,Molitoris2019}.
However, in prior research, the fixed effects and linear probability results do not 
deviate substantially.

@@

There has been substantial convergence in mortality risk across groups over time.
For intervals 12 months or longer, there is now little difference across 
the education groups, with even the no education group showing an infant mortality risk 
below five percent.

@@

% line 3661 in an_infant_mortality.log
The raw numbers for the group with the most uneven sex ratio indicates
that there is no impact on child mortality, even with high use of sex selection. 
A total of 1,004 women with twelve or more years of
education, who have no boys at the start of the third spell in the last
period, had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life, and half of those were 
born in the 0-23 months interval, and none in 48 months or more interval.

>>>>

The predicted parity progression probability is the likelihood of giving birth by the
imposed censoring based on standard survival curve calculations.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is to bias downward the differences in parity progression probabilities.
}
The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over, calculated as shown above.
For comparison, I also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
To find the expected average birth interval, I first calculate, for each woman, 
the probability of giving birth in each period $t$, which I use as weights to calculated 
her expected spell lengths. 
I then average the individual expected spell lengths across women using their parity 
progression probabilities as weights.
Finally, I add nine months to account for spell start to arrive at birth interval.

@@

In the latest period, the shortest average birth interval is 33 months for
women with no education in the second spell, an increase of a couple of
months from the first period.
The longest average interval was 51 months for the third spell for urban women with twelve or 
more years of education who had no boys, an increase of approximately a year.
For comparison, the median birth intervals reported for the NFHS have barely
moved over time, staying at approximately 31 months.%
\footnote{
See
\citet[][p.\ 110--112]{International-Institute-for-Population-Sciences-IIPS1995},
\citet[][p.\ 98--103]{International-Institute-for-Population-Sciences-IIPS2000},
\citet[][p.\ 88--91]{International-Institute-for-Population-Sciences-IIPS2007}, and
\citet[][p.\ 81--82]{International-Institute-for-Population-Sciences-IIPS2017}.
Median estimates are in
Appendix Tables \ref{tab:p25_p50_p75_low} through \ref{tab:p25_p50_p75_highest}.
}
This lack of change in the NFHS reported birth intervals underscores the importance of 
accounting for censoring of birth spells when trying to understand how birth spacing 
is changing over time.

Parity progression probabilities are lower the higher the education and are
falling across all groups over time, but the extent of the decline depends crucially on 
the sex composition of prior children and the parity.
Women with no sons are still the most likely to 
have the next birth, and this group has seen substantially smaller declines in
the likelihood of having another birth than women with one or more sons.%
\footnote{
However, there is also some evidence that parents prefer a mix of boys and girls: for 
the fourth spell women with two boys \emph{and} a girl
consistently have the lowest parity progression probability.
}
For urban women with eight to eleven years of education, for example, the second spell
probability of a next child declined by eight percentage points if the first child was a 
girl, but 15 percentage points if it was a boy.
Even more extreme is the fourth spell, where the likelihood of a birth declined by six 
percentage points to 75 percent if the woman had three children, but by almost 40 
percentage points to 25 percent if she had two boys and a girl.

@@

To better understand what factors are behind the changes in average spacing, I next 
examine its distribution---represented by the 25th and 75th percentiles 
of predicted birth intervals. 
This approach can capture differences across groups with similar changes in average spacing. 
For example, imagine that sex selection is behind an increase in average birth spacing, 
but that women still try to conceive early. 
In that case, we should see little change in the 25th percentile, but the 75th percentile 
spacing becomes much longer because those who use sex selection will have two or more 
pregnancies rather than one. 
In contrast, a general increase in spacing would show up as lengthening of both percentiles.

@@

I include the spell from marriage to first birth, despite the problems
capturing the exact timing of marriages since the estimated progression
probabilities should not be affected by this problem.
To ensure that the predictions depend solely on sample composition and the estimate, I
use the age at marriage for each woman and predict the likelihood of progressing to
each parity, assuming three years increases in age between each parity.
Shorter assumed increases in age between births leads to slightly higher predicted 
fertility.

@@

Despite the declines in the predicted cohort fertility, it is still mostly above the 
generally accepted replacement level of 2.1.
Only for urban women with twelve or more years of education is the predicted cohort
fertility clearly below 2.1, and even then, it is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Finally, the predicted cohort fertility numbers are likely too low,
especially for the lower education groups, since it relies only on the
first four births and because the model ignores births that took place past the end
of the imposed 105 month birth interval censoring.

@@

The dependent variable is whether the child died within the first twelve
months of life.
The main set of explanatory variables consists of dummies for the spacing from
the prior birth.
Starting nine months after the prior birth, the birth interval dummies cover 12 
month periods until 57 months, which covers until 105 months after the prior birth.
I use dummies for sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of index child, and the sex composition dummies
are all interacted.
Since the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

I estimate the probability of infant mortality using a Logit model.
Figures \ref{fig:mortality_low_med} and \ref{fig:mortality_high_highest} present 
the predicted probability of the second child dying within the first year at the 
possible combinations of index child sex, sex composition of prior children, and 
the birth spacing dummied.%
\footnote{
The online Appendix shows the corresponding graphs for the third child.
}
All other variables take their average values.
The graphs do not show confidence intervals to ease their legibility.

An important caveat is that the estimations do not 
address potential selection problems.
For example, if women who have difficulties conceiving or carrying a pregnancy 
to term also have a higher mortality risk for their offspring, a spurious correlation 
between long birth spacing and mortality may arise \citep{Kozuki2013}. 
Unfortunately, methods used previously to address selection, such
as family fixed effects, do not work well when the number of births is
low as it is for better-educated women \citep{Kozuki2013,Molitoris2019}.
However, the fixed effects and linear probability results did not 
deviate substantially in prior research.

@@

There has been substantial convergence in mortality risk across groups over time.
For intervals 21 months or longer, there is now little difference across 
the education groups, with even the no education group showing an infant mortality risk 
below five percent.

@@

% line 3661 in an_infant_mortality.log
The raw numbers for women with the most uneven sex ratio also suggests that even with 
high use of sex selection there is no impact on child mortality. 
A total of 1,004 women with twelve or more years of
education and no boys at the start of the third spell in the last
period, had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life.
Half of those who died were born in the 9-32 months interval, and none in 57 months 
or longer interval.
